biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-12-21 paper/PaperTRA/PaperTRA_rev1.tex
reduce do 6+1

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
A high proportion of car trips can be replaced by a combination of
public transit and cycling for the first-and-last mile. This paper
estimates the potential for cycling combined with public transit (PT) as
a substitute for car trips in the Lisbon metropolitan area and assesses
its socio-environmental impacts using open data and open source tools. A
decision support tool that facilitates the design and development of a
metropolitan cycling network was developed (\emph{biclaR}). The social
and environmental impacts were assessed using the \emph{HEAT for
Cycling} and the \emph{HEAT as a Service} tools. The impacts of shifting
car trips to PT were also estimated and monetized. The results indicate
that 10\% of trips could be made by bicycle + PT combination. Shifting
to cycling for the first-and-last mile stages can reduce annual
CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
10 years of at least €230 million. For the PT leg, the transfer from car
avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per year.
This evidence can support policymakers to prioritize interventions that
reduce the reliance on private motor vehicles.

\keywords{Active transport \and Intermodality \and First and last
mile \and Health economic assessment \and Environmental
impacts \and Open data and methods}

@@

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}4{]}, the
Lisbon's Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}5{]}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}6{]}. The
PCT has been extended to include cycling potential for travel to school
in England {[}7{]} and other trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel {[}8{]}. The OD Jittering method breaks down a
single point (i.e., the centroid of an area) into multiple random points
on the existing and neighboring road network, using OpenStreetMap as a
reference. This method then distributes the volume of trips within the
district among the randomly generated origin-destination pairs.

Using the
\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
package}, we employed a maximum disaggregation level of 100 trips per
O-D pair for this project. Figure \ref{fig:jitter} illustrates the
contrast between trip representation through the traditional method,
which connects a single desire line between each district, and the
presentation achieved through the randomization and disaggregation of
trips between districts, specifically for the Lisbon metropolitan area.

\begin{figure}

{\centering \includegraphics[width=1\linewidth,]{img/jitter}

\caption{Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

Although this method provides a more realistic representation of the
trips undertaken compared to the traditional approach, it does not fully
align with the actual O-D pairs of trips, which remain unknown due to
data privacy regulations.

@@

The mobility survey collects the origin and destination of trips but
does not include the respective routes. Modeling the realistic cycling +
PT routes between OD pairs depends on assumptions regarding the
characteristics of the cycling and road networks and the location of
public transport interfaces. Other constraints regarding the behavior of
potential cyclists determine the routing results. For example, such
restrictions can favor low speed, low traffic streets, more direct
routes, and less steep paths, among others, suitable for cycling.

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}9{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}10{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively. Different routing profiles enable
decision-makers to plan for different bicycle user typologies and/or for
different city cycling maturity levels {[}11{]}.

The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
metropolitan data aggregated and validated. This information is crucial
for an accurate PT trip and route estimation. A digital elevation model,
from the European Space Agency's COPERNICUS mission, was used to include
street gradient information, as a weight in cycling routing. The cycling
potential trips for the two national strategic targets (4\% and 10\%)
were estimated from the 2017 cycling and car trips (both as a driver and
as a passenger), the baseline scenario.

The routes were then overlaid and aggregated by segments, using
\href{https://docs.ropensci.org/stplanr/reference/overline.html}{\texttt{stplanr\ overline()}
R function}.

@@

The intermodality scenario considers trips combining PT and cycling for
the first and last legs. In a conservative approach, we have restricted
our analysis to the first and last legs with a combined length of up to
5 km (for instance: 1 km from origin to interface A plus 4 km from
interface B to destination) or up to 25 minutes on bike. Furthermore, we
have imposed restrictions on PT usage to include only trips with no PT
transfers, and up to 2 hours (120 min). Additionally, we have only
included PT modes that can easily accommodate bicycles, such as trains,
ferries, trams, and inter-municipal bus lines equipped with bike racks
(Figure \ref{fig:map1}).

\begin{figure}

@@

Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
the main PT interfaces in the LMA.

\begin{figure}

@@

For the cycling legs of the journey (first and last legs),
socio-environmental impacts were estimated, using the HEAT for Cycling
tool v5.0 {[}12{]} from the World Health Organization, and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
HEAT tool provided estimates on the shifting from car to cycling for a
short term time horizon (i.e., one year) and the long term (i.e., ten
years). We considered two dimensions: \emph{social} --- including the
physical activity of cyclists, air pollution exposure, and road
casualties; \emph{environmental} --- including CO\textsubscript{2}eq
emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values {[}13{]} for a Tier 3 detail level. We used
a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
trips were considered to be made under urban conditions and at an
average speed of 15 km/h during rush hour periods. Since the average
distance traveled per trip influences the overconsumption and emissions
from cold-start engine operation, we estimated energy and emission
factors for different ranges of trips at 500-meter intervals. In
addition, we assumed an occupancy rate of 1.6 passengers \emph{per} car
{[}3{]}.

Regarding PT, we considered the emission factor values reported in the
environmental and sustainability reports of the PT operators in the LMA
{[}14--17{]}.

Emissions were estimated for the following atmospheric pollutants: CO,
PM10, NOx, and VOC; and for the main green house gases:
CO\textsubscript{2}, CH\textsubscript{4}, and N\textsubscript{2}O,
converted in CO\textsubscript{2}eq. In particular, for the urban train
and tram -- with 100\% electric traction -- only CO\textsubscript{2}eq
emissions were considered (resulting from the production of electricity,
considering a ``well-to-tank'' approach), since the other pollutants are
not emitted locally.

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis {[}18{]} and the best up-to-date reference values for the
various gases {[}18--20{]}. The social impacts are in avoided premature
mortality. This result is finally monetized using the \emph{Statistical
Value of Life} for Portugal {[}21{]}. We updated all the monetary
reference values of the literature based on the annual inflation rate in
Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a discount rate of 5\% and inflation of
3\%.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT & Avoided CO2eq (ton/yr) & SE Benefits for 10 years (thousand €)\\
\midrule
4\% & safe & 538 514 & 2 312 & 20 385 & 5 920 & 230 270\\
4\% & direct & 500 880 & 2 274 & 18 944 & 6 011 & 223 720\\
10\% & safe & 538 514 & 2 312 & 52 324 & 15 192 & 591 790\\
10\% & direct & 500 880 & 2 274 & 48 609 & 15 414 & 574 200\\
\bottomrule
\end{tabular}
\end{table}

@@

Table \ref{tab:summary21} shows the potential trips by PT mode to
replace the second leg of the journey, in combination with cycling.
Trains offer the greatest potential for substitution (88\%). When
comparing the existing PT interfaces (Figure \ref{fig:map1}) with the
bike routes with highest potential to serve as first and last legs
(Figure \ref{fig:map2}) it becomes clear that the Train interfaces are
the ones that have the highest potential to attract car-to-PT
substituting trips, if their accessibility by bicycle is improved to be
safe.

\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
\toprule
Target & Routing & Potential & Bus & Ferry & Train & Tram\\
\midrule
4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{tab:summary22} presents the results of the avoided emissions
and its monetization for the second leg of the journey, by replacing car
trips with potential TP trips. Regarding the PT segment, the shift from
private car usage would lead to the mitigation of CO\textsubscript{2}
equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
million to €3.5 million yearly.

\caption{\label{tab:summary22}\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.}
\centering
\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}r}
\toprule
Target & Routing & CO2eq & CO & PM10 & NOx & VOC & Value (k€)\\
\midrule
4\% & safe & 8 593 & 17 & 1.9 & 27 & 0.8 & 1 425\\
4\% & direct & 8 702 & 18 & 2.0 & 28 & 0.8 & 1 453\\
10\% & safe & 20 627 & 42 & 4.6 & 65 & 2.0 & 3 431\\
10\% & direct & 20 793 & 42 & 4.7 & 66 & 1.9 & 3 487\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:summaryall}\label{summaryall}Summary of the avoided CO2eq emmissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.}
\centering
\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{8em}>{\raggedleft\arraybackslash}p{8em}>{\raggedleft\arraybackslash}p{8em}}
\toprule
Target & Routing & Avoided CO2eq (tons) & SE Benefits 1yr (k€) & SE Benefits 10yrs (k€)\\
\midrule
4\% & safe & 14 513 & 27 299 & 242 024\\
4\% & direct & 14 713 & 26 572 & 235 706\\
10\% & safe & 35 819 & 69 938 & 620 094\\
10\% & direct & 36 207 & 67 956 & 602 972\\
\bottomrule
\end{tabular}
\end{table}

Shifting from car to cycling + PT can reduce annual
CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year, and
the 10-year socio-environmental benefits account for €235 million to
€620 million, depending on the cycling targets.

The social impacts represent 98\% of the socio-environmental benefits
(in value) from replacing car trips to bicycle in first-and-last legs.
For the PT segment, we did not estimate the social impacts from
substituting car trips, although its health benefits would not be as
high as shifting to cycling.

The emissions of CO\textsubscript{2}eq that are avoided during both the
initial and final journey segments account for about 75\% of the
emissions avoided during the PT segment. This finding, while expected --
due the zero cycling emissions, should not be overlooked when promoting
the PT use. Improving the safe accessibility to PT interfaces to
cyclists and providing bicycle-friendly amenities such as parking
facilities can potentially lead to a higher reduction in
CO\textsubscript{2}eq emissions, compared to a scenario where
individuals shift from car travel to car + PT combination.

Our findings suggest that cycling and PT \emph{in combination} could
viably replace 10\% of current LMA trips, with an additional 6\% of PT
journeys prone to further substitution.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
conselho de ministros n.º 131/2019,
\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
online system for sustainable transport planning. Journal of Transport
and Land Use. 10, (2017).
https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}

\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
to school in england, and associated health and carbon impacts:
Application of the {`}propensity to cycle tool{'}. Journal of Transport
\& Health. 12, 263--278 (2019).
https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
jittering and routing options for converting origin-destination data
into route networks: Towards accurate estimates of movement at the
street level. The International Archives of the Photogrammetry, Remote
Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
(2022).
https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}

\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
cyclists: Review of market segmentation methods for planning practice.
Transportation Research Record. 2662, 125--133 (2017).
https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
Lieb, C., Oja, P., others:
\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
economic assessment tool ({HEAT}) for walking and for cycling: Methods
and user guide on physical activity, air pollution, injuries and carbon
impact assessments}. (2017).}

\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
emission inventory guidebook 2019,
\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Carris:
\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
de sustentabilidade 2019 - demostração não financeira}. {Carris -
Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Metropolitano de Lisboa:
\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{CP:
\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Transtejo:
\href{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}{Relatório
de sustentabilidade 2014}. {Grupo Transtejo, S.A.} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
Sirtori, E., Vignetti, S., Del Bo, C.:
\href{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}{Guide
to cost-benefit analysis of investment projects. Economic apraisal tool
for cohesion policy 2014-2020}. {European Commission - Directorate
General for Regional and Urban Policy} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
Hunt, A., Jong, G.D., Laird, J.:
\href{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}{{HEATCO
- Developing Harmonised European Approaches for Transport Costing and
Project Assessment. Deliverable 5, Proposal for Harmonised Guidelines.}}
(2006).}

\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
\CSLLeftMargin{20. }%
\CSLRightInline{Nash, C., others:
\href{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}{UNIfication
of accounts and marginal costs for transport efficiency: Final report}.
{Institute for Transport Studies, University of Leeds} (2003).}

>>>>

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This paper estimates the potential for cycling combined with public
transit (PT) as a substitute for car trips in the Lisbon metropolitan
area and assesses its socio-environmental impacts using open data and
open source tools. A decision support tool that facilitates the design
and development of a metropolitan cycling network was developed
(\emph{biclaR}). The social and environmental impacts were assessed
using the \emph{HEAT for Cycling} and the \emph{HEAT as a Service}
tools. The impacts of shifting car trips to PT were also estimated and
monetized. The results indicate that 10\% of trips could be made by
bicycle + PT combination. Shifting to cycling for the first-and-last
mile stages can reduce annual CO\textsubscript{2}eq emissions from 3,000
tons/day, with benefits over 10 years of at least €125 million. For the
PT leg, the transfer from car avoids of up to 20,500 tons of
CO\textsubscript{2}eq emissions per year. This evidence can support
policymakers to prioritize interventions that reduce the reliance on
private motor vehicles.

\keywords{Active transport \and Intermodality \and First and last
mile \and Health economic assessment \and Environmental
impacts \and Open methods}

@@

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively), the Lisbon's
Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}4{]}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}5{]}.
However, to the best of our knowledge, this is the first time that the
method has been integrated with public transport data using multi-modal
routing to estimate the potential and benefits of multi-stage cycling
and PT trips.

@@

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. The OD Jittering method breaks
down a single point (i.e., the centroid of an area) into multiple random
points on the existing and neighboring road network, using OpenStreetMap
as a reference {[}6{]}. Using the
\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
package}, we employed a maximum disaggregation level of 100 trips per
O-D pair for this project. Although this method provides a more
realistic representation of the trips undertaken compared to the
traditional approach, it does not fully align with the actual O-D pairs
of trips, which remain unknown due to data privacy regulations.

@@

The mobility survey collects the origin and destination of trips but
does not include the respective routes. Modeling the realistic cycling +
PT routes between OD pairs depends on assumptions regarding the
characteristics of the cycling and road networks and the location of
public transport interfaces.

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
which allows for great flexibility in configuring estimated route types.
\texttt{r5r} can calculate multi-modal routes using PT combined with
other modes. It enables the identification of the most direct or safest
cycling routes, using the Level of Traffic Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale. The routes were estimated for the base scenario for both
types of networks: \emph{direct} and \emph{safe}, using LTS 4 and LTS 3,
respectively.

The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
metropolitan data aggregated and validated. This information is crucial
for an accurate PT trip and route estimation.

@@

The intermodality scenario considers trips combining PT and cycling for
the first and last legs. In a conservative approach, we have restricted
our analysis to the first and last legs with a combined length of up to
5 km (for instance: 1 km from origin to interface A plus 4 km from
interface B to destination) or up to 25 minutes on bike. Furthermore, we
have imposed restrictions on PT usage to include only trips with no PT
transfers, and up to 2 hours (120 min). Additionally, we have only
included PT modes that can easily accommodate bicycles, such as trains,
ferries, trams, and inter-municipal bus lines equipped with bike racks
(Figure \ref{fig:map1}). Figure \ref{fig:map2} illustrates the resulting
bicycle routes to access the main PT interfaces in the LMA.

\begin{figure}[h]

@@

\begin{figure}[h]

@@

For the cycling legs of the journey (first and last legs),
socio-environmental impacts were estimated on the shifting from car to
cycling, using the HEAT for Cycling tool v5.0 {[}8{]} from the World
Health Organization, and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. We
considered two dimensions: \emph{social} --- including the physical
activity of cyclists, air pollution exposure, and road casualties;
\emph{environmental} --- including CO\textsubscript{2}eq emissions and
other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values {[}9{]} for a Tier 3 detail level.
Regarding PT, we considered the emission factor values reported in the
environmental and sustainability reports of the PT operators in the LMA.
The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis {[}10{]} and the best up-to-date reference values for the
various gases.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT & Avoided CO2eq (ton/yr) & SE Benefits for 10 years (thousand €)\\
\midrule
4\% & safe & 538 514 & 2 312 & 20 385 & 2 958 & 115 780\\
4\% & direct & 500 880 & 2 274 & 18 944 & 3 004 & 113 030\\
10\% & safe & 538 514 & 2 312 & 52 323 & 7 590 & 297 510\\
10\% & direct & 500 880 & 2 274 & 48 609 & 7 694 & 289 290\\
\bottomrule
\end{tabular}
\end{table}

@@

Regarding the PT mode to replace the second leg of the journey, in
combination with cycling, trains offer the greatest potential for
substitution (88\%). When comparing the existing PT interfaces (Figure
\ref{fig:map1}) with the bike routes with highest potential to serve as
first and last legs (Figure \ref{fig:map2}) it becomes clear that the
Train interfaces are the ones that have the highest potential to attract
car-to-PT substituting trips, if their accessibility by bicycle is
improved to be safe.

Regarding the PT segment, the shift from private car usage would lead to
the mitigation of CO\textsubscript{2} equivalent emissions to 8,500 to
20,800 tons annually, valued in €1.4 million to €3.5 million yearly.

@@

\caption{\label{tab:summaryall}\label{summaryall}Summary of the avoided CO2eq emmissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.}
\centering
\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{8em}>{\raggedleft\arraybackslash}p{8em}>{\raggedleft\arraybackslash}p{8em}}
\toprule
Target & Routing & Avoided CO2eq (tons) & SE Benefits 1yr (k€) & SE Benefits 10yrs (k€)\\
\midrule
4\% & safe & 11 551 & 14 380 & 127 534\\
4\% & direct & 11 706 & 14 135 & 125 016\\
10\% & safe & 28 217 & 36 861 & 325 814\\
10\% & direct & 28 487 & 35 905 & 318 062\\
\bottomrule
\end{tabular}
\end{table}

Shifting from car to cycling + PT can reduce annual
CO\textsubscript{2}eq emissions emissions by 11,500 to 28,500 tons per
year, and the 10-year socio-environmental benefits account for €125
million to €325 million, depending on the cycling targets

The environmental impacts represent less than 2\% of the
socio-environmental benefits (in value) from replacing car trips to
bicycle in first-and-last legs. For the PT segment, we did not estimate
the social impacts from substituting car trips, although its health
benefits would not be as high as shifting to cycling.

The emissions of CO\textsubscript{2}eq that are avoided during both the
initial and final journey segments account for about 74\% of the
emissions avoided during the PT segment. This finding, while expected --
due the zero cycling emissions, should not be overlooked when promoting
the PT use. Improving the safe accessibility to PT interfaces to
cyclists and providing bicycle-friendly amenities such as parking
facilities can potentially lead to a higher reduction in
CO\textsubscript{2}eq emissions, compared to a scenario where
individuals shift from car travel to car + PT combination.

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions.

@@

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
online system for sustainable transport planning. Journal of Transport
and Land Use. 10, (2017).
https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
Lieb, C., Oja, P., others:
\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
economic assessment tool ({HEAT}) for walking and for cycling: Methods
and user guide on physical activity, air pollution, injuries and carbon
impact assessments}. (2017).}

\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
emission inventory guidebook 2019,
\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
Sirtori, E., Vignetti, S., Del Bo, C.:
\href{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}{Guide
to cost-benefit analysis of investment projects. Economic apraisal tool
for cohesion policy 2014-2020}. {European Commission - Directorate
General for Regional and Urban Policy} (2014).}
